---
title: "Investigating Carbon Trends in the Atmosphere From 1960-Present"
author: "Sri-Amirthan Theivendran"
output:
  pdf_document: default
  html_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Introduction

Based on daily measurments of flask $\text{CO}_{2}$ taken from the Mauna Loa Observatory (MLO) in Hawaii at elevation $3397$ m from 1960-present, we investigate several hypotheses. First, although, carbon in the atmosphere is still increasing as shown the graphs below, we will investigate the claim that the increase in carbon has slowed recently. Also we will see whether there is a reasonable chance that carbon exceeds $430$ ppm by the year $2025$, which month results in most carbon being produced and whether the data supports the slowing of carbon emissions during global recessions and post-collapse of the Soviet Union. 
```{r, fig.align='center', fig.width= 8, fig.height=3, warning=FALSE, echo=FALSE, fig.cap="Recorded ${CO}_{2}$"}
#install.packages("Pmisc", repos="http://R-Forge.R-project.org")
cUrl=paste0("http://scrippsco2.ucsd.edu/assets/data/atmospheric/", "stations/flask_co2/daily/daily_flask_co2_mlo.csv")
cFile=basename(cUrl)
if(!file.exists(cFile)) download.file(cUrl, cFile)
co2s=read.table(cFile, header=FALSE, sep=",", skip=69, 
                stringsAsFactors = FALSE, 
                col.names=c("day", "time", "junk1", "junk2", "Nflasks", "quality", "co2"))
co2s$date=strptime(paste(co2s$day, co2s$time), format= "%Y-%m-%d %H:%M", tz="UTC")
co2s[co2s$quality>2, "co2"] = NA
par(mfrow=c(1,2))
plot(co2s$date, co2s$co2, cex=0.3,col="grey", xlab="time", ylab="ppm", main="Recorded CO2 Over Time")

plot(co2s[co2s$date>ISOdate(2015, 3, 1, tz="UTC"), c("date", "co2")], 
     type="o", xlab="time", ylab="ppm", cex=0.5, main="Recorded CO2 (Post March 2015)")
par(mfrow=c(1,1))
```

#### Methods

To the above data, we fit a GAM model. Specifically, we assume that the daily measurements $Y_i\sim \text{Gamma}(\mu_i/\nu, \nu)$ where the first parameter is the scale parameter and the second is the shape parameter. Our assumption is based on the following histogram of $\text{CO}_{2}$ measurements which look roughly gamma distributed. In any case, we use a log-linear model with covariates month, time (relative to $1980$), with the log-number of days as an offset, so that $\log \mu_i=X_i'\beta + f(\text{time}_i)$ where $X_i$ is the vector of covariates and $f$ is a cubic spline polynomial. We choose to include month in addition to a smoothly varying time trend in order to account for the apparent seasonality in the data. Obviously, a log-offset is needed to account for the varying numbers of days in months.
```{r, fig.align='center', fig.width= 8, fig.height=3, warning=FALSE, echo=FALSE, fig.cap="Histogram of ${CO}_{2}$ Measurements"}
timeOrigin=ISOdate(1980, 1, 1, 0, 0, 0, tz="UTC")
co2s$days=as.numeric(difftime(co2s$date, timeOrigin, units="days"))
co2s$month=as.factor(month.name[as.integer(strftime(co2s$date, "%m"))])
co2s$daysinMonth=Hmisc::monthDays(co2s$date)
co2s$nDays=log(co2s$daysinMonth)
hist(co2s$co2, xlab="ppm", main="Histogram of CO2 Measurements")
```

#### Results
The parameter estimates for the month coeffients on the natural scale are given below with the baseline being April. We see that carbon tends to higher in March than October. Carbon is highest during Februrary and lowest during October. A graph of the relative rate for the smoothly varying time trend (relative to baseline 1980) along with $95$ percent confidence bands is given below. It demonstrates that on the whole carbon in the atmosphere has been increasing for the last 40 years.
```{r, fig.align='center', fig.width= 8, fig.height=3, warning=FALSE, echo=FALSE}
co2gam=mgcv::gam(co2 ~ month+offset(nDays)+s(days), data=co2s, family = Gamma(link=log))
coefftable=summary(co2gam)$p.table[, 1:2]
paramest=round(exp(coefftable%*%Pmisc::ciMat(0.95))[-1, ], 3)
knitr::kable(paramest, caption="Relative rate for Each Month (Baseline is April)")
```

```{r, fig.align='center', fig.width= 8, fig.height=3, warning=FALSE, echo=FALSE, fig.cap="Relative Rate of Smooth Function"}
co2Pred=as.matrix(as.data.frame(mgcv::predict.gam(co2gam, co2s, type="terms", terms="s(days)", se.fit=TRUE)))
co2Pred=exp(co2Pred %*% Pmisc::ciMat())
#confidence bound of the relative rate for the time trend
matplot(co2s$days, co2Pred, log="y", xaxt="n", xlab="date", ylab="rr", col="black", lty=c(1,1,2), main="Relative Rate of Smooth Function")
pseq=seq(from=min(co2s$date), by="5 years", length.out = 15)
axis(1, at=difftime(pseq, timeOrigin, units="days"), labels=format(pseq, "%Y") )
```

#### Discussion
Now we turn our attention to forecasting. Based on the graph of recent $\text{CO}_2$ and predicted $\text{CO}_2$ emissions in the immediate future, we see that carbon will not comfortably exceed 430 ppm until roughly late $2027$. Ostenisbly, There is little chance that carbon will exceed 430 ppm by $2025$ based on the 95 percent confidence bands. However, we must take this conclusion with a grain of salt as the point confidence intervals glued together do not necessarily construct a confidence interval for the entire function.


```{r, fig.align='center', fig.width= 8, fig.height=3, warning=FALSE, echo=FALSE, fig.cap="Forecasted Emissions"}
newX=data.frame(date=seq(from=ISOdate(2020, 1, 1, tz="UTC"), by="months", length.out=12*8))
newX$days=as.numeric(difftime(newX$date, timeOrigin, units="days"))
newX$daysInMonth=Hmisc::monthDays(newX$date)
newX$nDays=log(newX$daysInMonth)
newX$month=as.factor(months(newX$date))

co2for=predict(co2gam, newX, se.fit=TRUE)
co2for=cbind(newX, co2for)
co2for$lower=co2for$fit-2*co2for$se.fit
co2for$upper=co2for$fit+2*co2for$se.fit
for (D in c("fit", "lower", "upper")){
  co2for[[paste(D, "exp", sep="")]]=exp(co2for[[D]])
}
plot(co2for$date, co2for$fitexp, type="n", xlab="date", ylab="ppm", ylim=c(400, 450), main="Forecasted C02 Levels")
matlines(co2for$date, co2for[, c("lowerexp", "upperexp", "fitexp")], lty=1, col=c("grey", "grey", "black"), lwd=c(2,2,1))
```


It is well-known that carbon levels have been rising in the past 40 years or so (and this observation is consistent with our model) but the rate at which carbon is increasing can fluctuate based on a variety of political and economic factors. Indeed, an approximation to the derivative of the smoothly varying function (and which represents the rate of carbon increase) is given below.

Around 2008, the graph shows that the rate of carbon increase was decreasing which coincides roughly with the global economic recession at this time. However, around 1980, the rate of carbon increase was increasing in spite of a global economic recession at that time. We see that in the immediate aftermath of the collapse of  the Soviet Union in 189, the rate  of carbon increase decreased and the rate of carbon increase reached a minimum around $1993$.

Finally, although carbon in the atmosphere is still increasing, the rate of carbon increase has plateaued quite recently.

```{r, fig.align='center', fig.width= 8, fig.height=3, warning=FALSE, echo=FALSE, fig.cap="Derivative of Smoothly Varying Function"}
newX=data.frame(date=seq(from=timeOrigin, by="months", length.out=12*40))
newX$days=as.numeric(difftime(newX$date, timeOrigin, units="days"))
newX$daysInMonth=Hmisc::monthDays(newX$date)
newX$nDays=log(newX$daysInMonth)
newX$month=as.factor(months(newX$date))


#predicting derivatives.
# new data for prediction
# prediction of smoothed estimates at each unique year value
# with standard error    
B <- predict(co2gam,  newX, type="response", se.fit=TRUE)


# finite difference approach to derivatives following
# example from ?predict.gam

eps <- 0.5
X0 <- predict(co2gam, newX, type = 'lpmatrix')


newXeps_p <- newX
newXeps_p$days<- newXeps_p$days+eps

X1 <- predict(co2gam, newXeps_p, type = 'lpmatrix')


# finite difference approximation of first derivative
# the design matrix
Xp <- (X1 - X0) / eps

# first derivative
fd_d1 <-  Xp %*% coef(co2gam)
newX$deriv<-fd_d1

#Derivative of  the smoothly varying function
plot(newX$date, newX$deriv, xlab="date", ylab="ppm/day", ylim=c(min(newX$deriv), max(newX$deriv)), main="Predicted Change in CO2 levels over Time", type="l", col="red")
```

#### Appendix

```{r, fig.align='center', fig.width= 8, fig.height=4, warning=FALSE, echo=TRUE, eval=FALSE}
#setting up the data
cUrl=paste0("http://scrippsco2.ucsd.edu/assets/data/atmospheric/", 
            "stations/flask_co2/daily/daily_flask_co2_mlo.csv")
cFile=basename(cUrl)
if(!file.exists(cFile)) download.file(cUrl, cFile)
co2s=read.table(cFile, header=FALSE, sep=",", skip=69, 
                stringsAsFactors = FALSE, 
                col.names=c("day", "time", "junk1", "junk2", 
                            "Nflasks", "quality", "co2"))
co2s$date=strptime(paste(co2s$day, co2s$time), format= "%Y-%m-%d %H:%M", tz="UTC")
co2s[co2s$quality>2, "co2"] = NA

#plot of recent CO2 emissions
par(mfrow=c(1,2))
plot(co2s$date, co2s$co2, cex=0.3,col="grey", xlab="time", 
     ylab="ppm", main="Recorded CO2 Over Time")

plot(co2s[co2s$date>ISOdate(2015, 3, 1, tz="UTC"), c("date", "co2")], 
     type="o", xlab="time", ylab="ppm", cex=0.5, main="Recorded CO2 (Post March 2015)")
par(mfrow=c(1,1))

#histogram of CO2 measurements
timeOrigin=ISOdate(1980, 1, 1, 0, 0, 0, tz="UTC")
co2s$days=as.numeric(difftime(co2s$date, timeOrigin, units="days"))
co2s$month=as.factor(month.name[as.integer(strftime(co2s$date, "%m"))])
co2s$daysinMonth=Hmisc::monthDays(co2s$date)
co2s$nDays=log(co2s$daysinMonth)
hist(co2s$co2, xlab="ppm", main="Histogram of CO2 Measurements")

#fitting the model and table of coefficients
co2gam=mgcv::gam(co2 ~ month+offset(nDays)+s(days), data=co2s, family = Gamma(link=log))
coefftable=summary(co2gam)$p.table[, 1:2]
paramest=round(exp(coefftable%*%Pmisc::ciMat(0.95))[-1, ], 3)
knitr::kable(paramest, caption="Relative rate for Each Month (Baseline is April)")

#predicted relative rate of the smooth function
co2Pred=as.matrix(as.data.frame(mgcv::predict.gam(co2gam, co2s, type="terms", 
                                                  terms="s(days)", se.fit=TRUE)))
co2Pred=exp(co2Pred %*% Pmisc::ciMat())
#confidence bound of the relative rate for the time trend
matplot(co2s$days, co2Pred, log="y", xaxt="n", xlab="date", ylab="rr", col="black", 
        lty=c(1,1,2), main="Relative Rate of Smooth Function")
pseq=seq(from=min(co2s$date), by="5 years", length.out = 15)
axis(1, at=difftime(pseq, timeOrigin, units="days"), labels=format(pseq, "%Y") )

#forecasting CO2 emissions from 2020-2028
newX=data.frame(date=seq(from=ISOdate(2020, 1, 1, tz="UTC"), by="months", length.out=12*8))
newX$days=as.numeric(difftime(newX$date, timeOrigin, units="days"))
newX$daysInMonth=Hmisc::monthDays(newX$date)
newX$nDays=log(newX$daysInMonth)
newX$month=as.factor(months(newX$date))

co2for=predict(co2gam, newX, se.fit=TRUE)
co2for=cbind(newX, co2for)
co2for$lower=co2for$fit-2*co2for$se.fit
co2for$upper=co2for$fit+2*co2for$se.fit
for (D in c("fit", "lower", "upper")){
  co2for[[paste(D, "exp", sep="")]]=exp(co2for[[D]])
}
plot(co2for$date, co2for$fitexp, type="n", xlab="date", ylab="ppm", 
     ylim=c(400, 450), main="Forecasted C02 Levels")
matlines(co2for$date, co2for[, c("lowerexp", "upperexp", "fitexp")], 
         lty=1, col=c("grey", "grey", "black"), lwd=c(2,2,1))

#Predicting Derivatives

newX=data.frame(date=seq(from=timeOrigin, by="months", length.out=12*40))
newX$days=as.numeric(difftime(newX$date, timeOrigin, units="days"))
newX$daysInMonth=Hmisc::monthDays(newX$date)
newX$nDays=log(newX$daysInMonth)
newX$month=as.factor(months(newX$date))


#predicting derivatives.
# new data for prediction
# prediction of smoothed estimates at each unique year value
# with standard error    
B <- predict(co2gam,  newX, type="response", se.fit=TRUE)


# finite difference approach to derivatives following
# example from ?predict.gam

eps <- 0.5
X0 <- predict(co2gam, newX, type = 'lpmatrix')


newXeps_p <- newX
newXeps_p$days<- newXeps_p$days+eps

X1 <- predict(co2gam, newXeps_p, type = 'lpmatrix')


# finite difference approximation of first derivative
# the design matrix
Xp <- (X1 - X0) / eps

# first derivative
fd_d1 <-  Xp %*% coef(co2gam)
newX$deriv<-fd_d1

#Derivative of  the smoothly varying function
plot(newX$date, newX$deriv, xlab="date", ylab="ppm/day", 
     ylim=c(min(newX$deriv), max(newX$deriv)), 
     main="Change in CO2 levels over Time", type="l", col="red")
```

